---
title: "Lab 7 Assignment: Group 05"
format:
  html:
    embed-resources: true
editor: visual
---

# Load libraries

```{r}
#| message: false
library(tidyverse)
library(broom)
library(cowplot)
```

# Load data

We use the biopsy data from https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv and the gravier data from https://pubmed.ncbi.nlm.nih.gov/20842727/

```{r}
biopsy <- read_csv("/net/pupil1/home/people/s233427/projects/lab07_git_exercises/Data/biopsy.csv")
gravier <- read_tsv("/net/pupil1/home/people/s233427/projects/lab07_git_exercises/Data/02_gravier_clean.gz")
```

# PCA
We start by insepecting our data
```{r}
biopsy |> 
  sample_n(10)
```

```{r}
gravier |>
  select(y) |> 
  sample_n(10)

```


As the `prcomp()` function used for PCA only takes nummeric column as input. We start by removing all non-numeric attributes from our data. Futhermore, we normalize the data by scale to unit variance using `scale = TRUE`.

```{r}
pca_fit_biopsy <- biopsy |> 
  select(where(is.numeric)) |>  # Remove all non-numeric columns
  prcomp(scale = TRUE) # Perform PCA and scale to unit variance

pca_fit_gravier <- gravier |> 
  select(where(is.numeric)) |>  # Remove all non-numeric columns
  prcomp(scale = TRUE) # Perform PCA and scale to unit variance

```

## Visualisation of data in PC coordinates
We then use `augment()` from `broom` to fit the original data to our fitted model. The column cotaining the fitted coordinates are `.fittedPC1` and `.fittedPC2`. We further stratify by outcome. 

```{r}
pca_fit_biopsy |> 
  augment(biopsy) |>  # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", benign = "#0072B2")
  ) +
  theme_half_open(12) + background_grid()
```
```{r}
pca_fit_gravier |> 
  augment(gravier) |>  # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = y)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(poor = "#D55E00", good = "#0072B2")
  ) +
  theme_half_open(12) + background_grid()
```
## Visualization of attribute direction in 2D PC coordinates
```{r}
# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit_biopsy %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

